Verification in Isabelle/HOL of a Set-based Algorithm for Computing Strongly Connected Components

### 2022-09-13 cpp/submission.tex
termination proof

@@

@@

>>>>

\paragraph{Outline of the paper.}

Section~\ref{sec:background} introduces necessary background for our work, in particular the proof assistant Isabelle/HOL and our representation of finite directed graphs and strongly connected components. Our encoding of Bloemen's algorithm in Isabelle is described in Section~\ref{sec:formalization}. The presentation of the correctness proof appears in Sections~\ref{sec:partial-correctness}, introducing the main invariants and the pre- and post-conditions used in the proof of partial correctness, and~\ref{sec:termination}, which explains the proof of termination. An assessment of the proof effort appears in Section~\ref{sec:effort}, and Section~\ref{sec:conclusion} contains a summary of the work and some perspectives for future work.

@@

Informally, the algorithm terminates because (i) whenever function \prog{dfs} is called for arguments~$v$ and~$e$, the node~$v$ will be added to the set of \prog{visited} nodes (and therefore the complement of that set w.r.t.\ the finite set of vertices decreases) and (ii) every call of function $\prog{dfss} v e$ either terminates immediately or adds a new node $w$ to $\prog{vsuccs}~e~v$. Observe that this is true only if the calls respect the pre-conditions: for example, (i) does not hold if~$v$ was already visited at the call of \prog{dfs}. We now explain how we make this argument formal and prove termination of the algorithm, assuming the pre-conditions.

As a first step, we define the following relation on argument tuples of the two functions.\footnote{Remember that \prog{Inl} and \prog{Inr} inject arguments of function \prog{dfs}, respectively \prog{dfss}, into the sum type representing the pair of functions.}

\begin{small}
\begin{lstlisting}[language=isabelle]
definition dfs-dfss-term where dfs-dfss-term ==
  { (Inr (v,e1), Inl (v,e)) | v e e1.
    v sin vertices - visited e 
    /\ visited e1 = visited e cup {v} }
cup { (Inl (w,e), Inr (v,e)) | v w e. v sin vertices }
cup { (Inr (v,e''), Inr (v,e)) | v e e''.
    v sin vertices /\ sub-env e e''
    /\ (exists w sin vertices. w sin vsuccs e'' v - vsuccs e v) }
\end{lstlisting}
\end{small}

This relation over-approximates the pairs of argument tuples for which recursive calls occur. For example, the first set in the above definition represents calls of \prog{dfss} from \prog{dfs} where $v$ was not visited at the call of \prog{dfs} but is added to the set of visited nodes at the call of \prog{dfss}. Observe that the second set, which corresponds to the call of \prog{dfs} from \prog{dfss}, requires that the call environment remains unchanged and in particular does not impose an immediate increase of either set \prog{visited} or \prog{vsuccs}. However, the use of different injection functions implies that one cannot construct a cycle using these argument tuples.

Crucially, we prove that the relation \prog{dfs-dfss-term} is well-founded.

\begin{small}
\begin{lstlisting}[language=isabelle]
lemma wf-term: wf dfs-dfss-term
\end{lstlisting}
\end{small}

In the statement of the lemma, \prog{wf} represents the well-foundedness predicate, which is predefined in the Isabelle standard library. The lemma is proved by embedding the argument tuples into triples $(a,b,c)$ where $a$ is the set of unvisited vertices, $b$ is the set of edges between vertices that have not been followed, and $c$ is $0$ if the triple represents an argument of \prog{dfs} and $1$ if it is an argument of \prog{dfss}. These triples can be ordered by the strict subset order for the two first component and the predecessor relation on natural numbers for the third one, and Isabelle determines that the product of these orders is well-founded. Finally, \prog{dfs-dfss-term} can be embedded as an inverse image in this well-founded relation, and is therefore itself well-founded.

We now prove that the previously defined pre-conditions imply termination of our two functions.

\begin{small}
\begin{lstlisting}[language=isabelle]
theorem dfs-dfss-termination:
  [| v sin vertices; pre-dfs v e |] 
  ==> dfs-dfss-dom (Inl (v,e))
  [| v sin vertices; pre-dfss v e |] 
  ==> dfs-dfss-dom (Inr (v,e))
\end{lstlisting}
\end{small}

The proof proceeds by well-founded induction along the relation \prog{dfs-dfss-term} and then shows that the arguments of recursive calls are indeed related. It makes use of the introduction rule for the predicate \prog{dfs-dfss-dom} that was generated by Isabelle for the definition of the two mutually recursive functions.

Putting everything together, we can now prove our main theorem asserting the total correctness of the algorithm.

\begin{small}
\begin{lstlisting}[language=isabelle]
theorem correctness:
  assumes v sin vertices
  shows   sccs (dfs v (init-env v)) = 
          {S . is-scc S /\ (forall n sin S. reachable v n)}
\end{lstlisting}
\end{small}

\section{Proof Effort}
\label{sec:effort}

\textbf{Summarize the overall effort of the proof in number of lines (definitions / proofs) for each main category.}

@@

\begin{itemize}
\item successfully proved correctness of an efficient sequential SCC algorithm
\item compare to Tarjan proof (Chen et al.): both are DFS algorithms, the algorithm considered here collapses loops as soon as they are found, minimizing stack size
\item compare to Lammich's work: he exploits the refinement framework for deriving executable code, we work on a higher-level representation of the algorithm as elementary recursive functions, hence invariants are simpler
\item objective: stepping stone towards a proof of the parallel version of the algorithm from Bloemen's thesis -- understand which invariants can be relaxed.
\end{itemize}
